When two-point correlation function fails to describe a structure...

### 2024-02-29 paper.tex
Results

\begin{document}
\title{When two-point correlation function fails to describe a structure...}
\author{Foo Bar}
\maketitle

@@

To find non-trivial collisions we firstly introduce a change of variables
$x = z^{-1}$ which is justified by a symmetry of \cref{eq:s2z}. Secondly we
introduce a notion of a reverse of a polynomial and palindrome polynomials.
\begin{definition}
  A reverse of a polynomial $f(x)$ is a polynomial $x^{\deg f}f(x^{-1})$. A
  polynomial is called a palindrome if it is equal to its reverse.
\end{definition}
Examples of palindrome polynomials include $x$, $x + 1$, $x^2 - 2x + 1$, but not
$x^3+x+1$.

@@

A stronger bound of the probability of the same factorization with higher
reliance on the Riemann hypothesis is given by:
\begin{equation}
  p = 1 - \exp(-c \sqrt{d} / \log d)
  \label{eq:prob-cyc-2}
\end{equation}
if the Riemann hypothesis holds for $\zeta_k$ for $K = \mathbb{Q}$ and all
fields $K = \mathbb{Q}(a)$ where $a$ is a root of $f$.

Now we will prove a simple lemma which connects these results with collisions in
correlation functions.
\begin{lemma}
  Let $A$ be a sufficiently long random sequence of 0's and 1's where each
  element is chosen randomly and independently with equal probability $1/2$ with
  exception of the first and the last elements which are 1's. Then, with
  probability at least
  \begin{equation*}
    p = 1 - 2\exp(-c(\log d)^\beta)
  \end{equation*}
  for some $c > 0$, $\beta > 1$ the sequence $A$ of the length $d+1$ does not
  have non-trivial collisions in the two-point correlation function.
\end{lemma}
\begin{proof}
  Let $f(x)$ be an image of $A$ under Z-transform and a change of variables
  $x = z^{-1}$. With probability at least $p$ it factors in $\mathbb{Z}[x]$ as
  $\Phi \tilde{f}$, where $\Phi$ is a product of cyclotomic polynomials and
  $\tilde{f}$ is irreducible. All cyclotomic polynomials with exception of
  $\Phi_1 = x - 1$ are palindromes. $\Phi_1$ does not divide $f$ because
  $f(1) = \sum\limits_{k=0}^{d-1} f_k \ge 2$. Therefore, $f$ has at most one
  non-palindrome factor and hence $A$ does not have non-trivial collisions.
\end{proof}

For the multidimensional case, our bound is more pessimistic:
\begin{lemma}
  Let $A$ be a sufficiently big random $D-dimensional$ array of 0's and 1's
  where each element is chosen randomly and independently with equal probability
  $1/2$ with exception of the first and the last elements found on the edge
  $\mathcal{L}$, which are 1's. Then, with probability at least
  \begin{equation*}
    p = 1 - \sqrt{\frac{2}{\pi d}}
  \end{equation*}
  where $d+1$ is the length of $\mathcal{L}$, $A$ does not have non-trivial
  collisions in the two-point correlation function.
\end{lemma}
\begin{proof}
  Firstly, assume $D = 2$. Let $f(x_1, x_2) \in \mathbb{Z}[x_1][x_2]$ be an
  image of $A$ under Z-transform and change of variables $x_i = z_i^{-1}$. Let
  $\phi_l$ be a mapping $\mathbb{Z}[x_1][x_2] \rightarrow \mathbb{Z}[x_1]$ which
  ``extracts'' the leading coefficient $f_l \in \mathbb{Z}[x_1]$ from a
  bivariate polynomial $f$ and $\phi_f$ be a mapping
  $\mathbb{Z}[x_1][x_2] \rightarrow \mathbb{Z}[x_1]$ which ``extracts'' the free
  coefficient $f_0 \in \mathbb{Z}[x_1]$.

  $\phi_f$ and $\phi_l$ are homomorphisms from multiplicative monoid
  $\mathbb{Z}^*[x_1][x_2]$ to $\mathbb{Z}^*[x_1]$.
  Indeed if $g = g_0 + g_1x_2 + \dots + g_nx_2^n$ and
  $h = h_0 + h_1x_2 + \dots + h_mx_2^m$ then
  \begin{align*}
    \phi_f(g h) &= \phi_f[(g_0  h_0) + (g_0 h_1 + g_1 h_0)x_2 + \dots + (g_n
      h_m) x_2^{n+m}] \\
    &= g_0 h_0 = \phi_f(g) \phi_f(h) \qquad \text{and} \\
    \phi_l(g h) &= \phi_l[(g_0  h_0) + (g_0 h_1 + g_1 h_0)x_2 + \dots + (g_n
      h_m) x_2^{n+m}] \\
    &= g_n h_m = \phi_l(g) \phi_l(h)
  \end{align*}

  Now, without loss of generality, assume that the free (resp. leading)
  coefficient $f_k$ of $f$ has a degree $d$ and a non-zero free coefficient. If
  it is irreducible in $\mathbb{Z}[x_1]$, then $f$ is irreducible in
  $\mathbb{Z}[x_1][x_2]$. Indeed, if $f$ is not irreducible, then $f = p_1 p_2$
  for some non-constant $p_1, p_2$. Applying $\phi = \phi_f$ (resp. $\phi_l$) to
  this equality, we get $\phi(f) = f_k = \phi(p_1) \phi(p_2)$ for non-constant
  $\phi(p_1), \phi(p_2)$ which contradicts with $f_k$ being irreducible. Using
  \cref{eq:prob-irr} we obtain the least bound for the probability of $f$ being
  irreducible. If $f$ is irreducible, then it cannot have two or more
  non-palindrome factors and therefore cannot have non-trivial collisions, which
  proves the lemma.

  The proof for $D > 2$ is by recursion.
\end{proof}

@@

\section{Results}

>>>>

\begin{document}
\title{When two-point correlation function fails to describe a structure...}
\author{Biba and Boba}
\maketitle

@@

To find non-trivial collisions we firstly introduce a change of variables
$x = z^{-1}$ which is justified by a symmetry of \cref{eq:s2z}. Secondly we
introduce a notion of a reverse \textcolor{red}{RECIPROCAL!} of a polynomial and palindrome polynomials.
\begin{definition}
  A reverse of a polynomial $f(x)$ is a polynomial $x^{\deg f}f(x^{-1})$. A
  polynomial is called a palindrome if it is equal to its reverse.
\end{definition}
Examples of palindrome polynomials include $x$, $x + 1$, $x^2 - 2x + 1$, but not
$x^3+x+1$.

@@

Now we will prove a simple lemma which connects these results with collisions in
correlation functions.
\begin{theorem}
  \label{the:sequence}
  Let $A$ be a sufficiently long random sequence of 0's and 1's where each
  element is chosen randomly and independently with equal probability $1/2$ with
  exception of the first and the last elements which are 1's. Then, with
  probability at least
  \begin{equation*}
    p = 1 - 2\exp(-c(\log d)^\beta)
  \end{equation*}
  for some $c > 0$, $\beta > 1$ the sequence $A$ of the length $d+1$ does not
  have non-trivial collisions in the two-point correlation function.
\end{theorem}
\begin{proof}
  Let $f(x)$ be an image of $A$ under Z-transform and a change of variables
  $x = z^{-1}$. With probability at least $p$ it factors in $\mathbb{Z}[x]$ as
  $\Phi \tilde{f}$, where $\Phi$ is a product of cyclotomic polynomials and
  $\tilde{f}$ is irreducible. All cyclotomic polynomials with exception of
  $\Phi_1 = x - 1$ are palindromes. $\Phi_1$ does not divide $f$ because
  $f(1) = \sum\limits_{k=0}^{d-1} f_k \ge 2$. Therefore, $f$ has at most one
  non-palindrome factor and hence $A$ does not have non-trivial collisions.
\end{proof}

To obtain some results for the multidimensional case, we need to prove an
auxiliary result first.
\begin{lemma}
  \label{lem:factors}
  Let $\phi$ be a homomorphism from $R[x]$ to $R'[x]$ and $\phi f \ne const$ if
  $f \ne const$. When $f$ has no more irreducible factors (counted with their
  repetitions) than $\phi f$.
\end{lemma}
\begin{proof}
  Suppose $f = \prod_{k=1}^n f_k^{m_k}$ where $f_k$ is irreducible in
  $R[x]$. Applying $\phi$ to $f$ we get $\phi f = \prod_{k=1}^n (\phi f_k)^{m_k}$.
  Since no $\phi f_k$ is constant, their product cannot be contracted to an
  irreducible polynomial in $R'[x]$. Each factor $\phi f_k$ can be factored to
  one or more irreducible polynomials in $R'[x]$ which proves the lemma.
\end{proof}

Now we get our main result:
\begin{theorem}
  Let $A$ be a 2-dimensional $D_1 \times D_2$ array of 0's and 1's where each
  element is chosen randomly and independently with probability $1/2$ with
  exception of three elements $A[0, 0]$, $A[D_1 - 1, 0]$ and $A[0, D_2 - 1]$,
  which are 1's. Then $A$ does not have non-trivial collisions in the two-point
  correlation function with probability $p \to 1$ as $D_1, D_2 \to \infty$.
\end{theorem}
\begin{proof}
  Let $f(x, y)$ be an image of $A$ under Z-transform and a change of variables
  $x = z_1^{-1}, y = z_2^{-1}$. Let $\phi f = f \mod y$ be a
  ``partial evaluation'' homomorphism from $\mathbb{Z}[x, y]$ to
  $\mathbb{Z}[x]$. Let $f_x(x) = \phi f = f(x,0)$. Under conditions of the
  theorem $f(0) = 1$ and $d = \deg f_x = D_2 - 1$.  According to
  \cref{eq:prob-irr} we have $P\{\text{$f_x$ is irreducible}\} \to 1$ with
  $d \to \infty$.

  We have three possible cases of factorization of $f$:
  \begin{itemize}
  \item $f$ is irreducible.
  \item $f$ factors as $f(x, y) = g(x)h(y)$. In this case
    $\phi h = h(0) = const$ and \cref{lem:factors} cannot be applied. According
    to distribution of coefficients in $f$ the probability of this case is
    $P_{split} = 2^{D_1 + D_2 - D_1 D_2 - 1}$.
  \item $f$ factors as $f(x, y) = g(x, y) h(x, y)$. According to conditions of
    the theorem, $\phi g \ne const$ and $\phi h \ne const$. We have
    $f_x = (\phi g) (\phi h)$. As we have shown, the probability of this event
    tends to zero with $D_2 \to \infty$. Applying \cref{lem:factors}, we can say
    the same about probability of this case.
  \end{itemize}
  Hence, the probability of the first case tends to one, with increase in
  $D_1, D_2$. If $f$ is irreducible, there is no two or more non-palindrome
  factors of $f$, hence $A$ does not have non-trivial collisions.
\end{proof}
This theorem also can be proved for multidimensional arrays by recursively
applying the same argument.

@@

\section{Results}
We have written a program in Common Lisp language
(\url{https://github.com/fatimp/ac-collisions-mt}) which searches for collisions
in the two-point correlation functions for random sequences of 0's and 1's with
the first and the last elements being equal to 1. The number of collisions for
each length are present in \cref{tab:numbers}. Some examples of non-trivial
collisions are in \cref{tab:collisions}.
\begin{table}[!pt]
  \centering
  \begin{tabular}{|c|c|c||c|c|c|}
    \hline
    $d$ & $N_{col}$ & $N_{irred}$ & $d$ & $N_{col}$ & $N_{irred}$\\
    \hline
    \hline
    25 & 20040 & 15146508 & 38 & 344 & 16269320 \\
    \hline
    26 & 13698 & 15400557 & 39 & 278 & 16137897 \\
    \hline
    27 & 10787 & 15241401 & 40 & 173 & 16454421 \\
    \hline
    28 & 7615 & 15656751 & 41 & 128 & 16257667 \\
    \hline
    29 & 5808 & 15365485 & 42 & 89 & 16565040 \\
    \hline
    30 & 4199 & 15883361 & 43 & 62 & 16383822 \\
    \hline
    31 & 3285 & 15595498 & 44 & 39 & 16634958 \\
    \hline
    32 & 2315 & 16023603 & 45 & 39 & 16511214 \\
    \hline
    33 & 1782 & 15821891 & 46 & 23 & 16630491 \\
    \hline
    34 & 1161 & 15988106 & 47 & 34 & 16488227 \\ 
    \hline
    35 & 926 & 15811952 & 48 & 20 & 16860506 \\
    \hline
    36 & 589 & 16356337 & 49 & 9 & 16691221 \\
    \hline
    37 & 471 & 16102018 & 50 & 13 & 16829791 \\
    \hline
  \end{tabular}
  \caption{The number of non-trivial collisions in the two-point correlation
    function and the number of irreducible polynomials for degrees
    $d = [25 \dots 50]$. The degree $d$ equals to the length of a random
    sequence minus one. The collisions were calculated by making $2 \cdot 10^7$
    random trials for each degree (polynomials were taken with repetitions).}
  \label{tab:numbers}
\end{table}
\begin{table}[!pt]
  \centering
  \begin{tabular}{|c|r|}
    \hline
    $d$ & Sequences which are non-trivial collisions \\
    \hline
    25 & 10000001000100000000100101 \\
    & 10000100010010000000000101 \\
    \hline
    & 10111111011111101111110011 \\
    & 11111011111101111110111001 \\
    \hline
    30 & 1000010000010100000000000001001 \\
    & 1000000010000010000000000101001 \\
    \hline
    & 1110111101010001100101100001111 \\
    & 1111110101001011001000111000111 \\
    & 1111010010000111000110111010111 \\
    & 1110011100001001010111010011111 \\
    \hline
    & 1110111111010101011011110111111 \\
    & 1111110111010110101011111110111 \\
    \hline
    & 1001001011011110110100110100001 \\
    & 1000010010110110110101111001001 \\
    \hline
    40 & 10010110000000000000000010011011001001011 \\
    & 11010010000000000000000011011001001101001 \\
    \hline
    & 10111111010101111111101010011111111100011 \\
    & 10010101111111101010111111011100011111111 \\
    \hline
    & 11011010010011101100011100101100110100001 \\
    & 10010010110111001000111101001100101100011 \\
    \hline
  \end{tabular}
  \caption{Examples of non-trivial collisions in the two-point correlation
    function.}
  \label{tab:collisions}
\end{table}

We have experimentally computed the probability of a sequence not having
non-trivial collisions and compared it with results of \cref{the:sequence}. The
result of comparison (taking $c = 0.148, \beta = 3.373$) is on
\cref{fig:prob-comparison}.
\begin{figure}
  \centering
  \includegraphics[width=0.5\linewidth]{images/prob.png}
  \caption[]{Probability of a sequence of 0's and 1's with length $d+1$ and the
    first and last elements being equal to 1 not to have non-trivial collisions
    in the two-point correlation function: the esimate from \cref{the:sequence}
    and a computer experiment.}
  \label{fig:prob-comparison}
\end{figure}

For two-dimensional arrays we were able to find non-trivial collisions for array
sizes at most $7 \times 7$ (with $10^6$ trials). One of such collisions is on
\cref{fig:collision-2d}.
\begin{figure*}[tp]
  \centering
  \includegraphics[width=0.3\linewidth]{images/collision-2d-1.png}
  \hfill
  \includegraphics[width=0.3\linewidth]{images/collision-2d-2.png}
  \hfill
  \includegraphics[width=0.3\linewidth]{images/collision-2d-s2.png}
  \caption[]{Two 2-dimensional $7 \times 7$ arrays (left, middle), having the
    same two-point correlation function (right).}
  \label{fig:collision-2d}
\end{figure*}

\newpage
